---
title: "Assignment 1: The cocktail bar"
author: "Anna Oszvald"
date: "2023.12.12."
output: html_document
---


## 1. Read the data

```{r}
# 1. Read the data
cocktail_data <- readr::read_tsv('https://raw.githubusercontent.com/nthun/cocktail-balance/master/cocktail_data.tsv')
```

## 2. Transform the data table and clean the ingredient variable!


```{r}
#Overview
print(summary(cocktail_data))


cat("Column Headers:\n")
print(names(cocktail_data))


cat("\nFirst 2 Rows of the Data:\n")
print(head(cocktail_data, 2))


# Load packages
if (!requireNamespace("tidyverse", quietly = TRUE)) {
  # If not, install tidyverse
  install.packages("tidyverse", repos = "http://cran.rstudio.com/")
}

library(tidyverse)

# Put ingredients in separate rows.
cocktail_data <- cocktail_data %>%
  separate_rows(ingredients, sep = "<br>") %>%
  mutate(ingredients = str_trim(ingredients))

# Clean of all quantities (e.g., 1/2 oz, 2 dashes, etc.) from ingredients
cocktail_data <- cocktail_data %>%
  mutate(ingredients = str_replace_all(ingredients, "\\d+/?\\d*\\s*(oz|dashes|tsp|tbsp|bsp|dash|drops|drop)?", "")) %>%
  mutate(ingredients = str_trim(ingredients))

# Clean the indicator of alcohol content (e.g., 47.3% abv) from ingredients
cocktail_data <- cocktail_data %>%
  mutate(ingredients = str_replace_all(ingredients, "\\d+\\.?\\d*% abv", "")) %>%
  mutate(ingredients = str_replace_all(ingredients, "<br/>", ", ")) %>%
  mutate(ingredients = str_replace_all(ingredients, "\\(%", "")) %>%
  mutate(ingredients = str_replace_all(ingredients, "abv\\)", "")) %>%
  mutate(ingredients = str_replace_all(ingredients, "\\(.%", "")) %>%
  mutate(ingredients = str_replace_all(ingredients, "<B>", "B>,")) %>%
  mutate(ingredients = str_replace_all(ingredients, "^\\.", "")) %>%
  mutate(ingredients = str_replace_all(ingredients, fixed("cr\u008fme"), "creme")) %>%
  mutate(ingredients = str_replace_all(ingredients, fixed("simple"), "simple syrup")) %>%
  mutate(ingredients = str_replace_all(ingredients, fixed("Marschino"), "maraschino")) %>%
  mutate(ingredients = str_replace_all(ingredients, fixed("crem de violette"), "creme de violette")) %>%
  mutate(ingredients = str_replace_all(ingredients, fixed("sugard"), "sugared")) %>%
  mutate(ingredients = str_replace_all(ingredients, fixed("syrup syrup"), "syrup")) %>%
  mutate(ingredients = str_replace_all(ingredients, fixed("syrup syrup"), "syrup")) %>%
mutate(ingredients = str_replace_all(ingredients, fixed("cherry herring "), "herring cherry liqueur")) %>%
  mutate(ingredients = str_trim(ingredients))

# Separate "Luxardo Marschino<B> Plymouth Gin"
cocktail_data <- cocktail_data %>%
  mutate(ingredients = str_replace_all(ingredients, "<b> ", ",")) %>%
  separate_rows(ingredients, sep = ",")

# Remove extra spaces, ". ", and make all letters lowercase
cocktail_data <- cocktail_data %>%
  mutate(ingredients = str_trim(ingredients)) %>%
  mutate(ingredients = str_replace_all(ingredients, "\\s*\\.\\s*", "")) %>%
  mutate(ingredients = stringr::str_to_lower(ingredients))


# Get unique ingredients
unique_ingredients <- cocktail_data %>%
  distinct(ingredients) %>%
  arrange(ingredients) %>%
  mutate(ingredients = str_replace_all(ingredients, fixed("cherry herring"), "herring cherry liqueur")) %>%
  mutate(ingredients = str_replace_all(ingredients, fixed("lairds applejack bottled in bond"), "applejack apple brandy"))

# Remove the specified ingredients
unique_ingredients <- unique_ingredients[!grepl("bitters", unique_ingredients$ingredients), ]
unique_ingredients <- unique_ingredients[!grepl("lime juice or champagne acid", unique_ingredients$ingredients), ]

# Display unique ingredients
print("\nUnique Ingredients:")
print(unique_ingredients)

# Get unique ingredients
unique_ingredients <- cocktail_data %>%
  distinct(ingredients) %>%
  arrange(ingredients) %>%
  mutate(ingredients = str_replace_all(ingredients, fixed("cherry herring"), "herring cherry liqueur")) %>%
  mutate(ingredients = str_replace_all(ingredients, fixed("lairds applejack bottled in bond"), "applejack apple brandy"))

# Remove the specified ingredients
unique_ingredients <- unique_ingredients[!grepl("bitters", unique_ingredients$ingredients), ]
unique_ingredients <- unique_ingredients[!grepl("lime juice or champagne acid", unique_ingredients$ingredients), ]

# Display unique ingredients
print("\nUnique Ingredients:")
print(unique_ingredients)


```

## 3. All ingredients in alphabetical order

```{r}
## 3. All ingredients in alphabetical order

cat("\nUnique Ingredients:\n")
print(unique_ingredients)


```

## 4. Number of unique ingredients

```{r}
# Display unique ingredients
cat("\nUnique Ingredients:\n")
print(unique_ingredients, n = Inf)

#You would need 59 unique ingredients, if you consider different branded and differently processed products as unique ingredient.
```

```{r}
# Example data frame with ingredients
ingredients <- c("absinthe", "absolut citron vodka", "agar-clarified grapefruit juice",
                 "amer picon", "angostura bitters", "apple brandy", "benedictine",
                 "blanco tequila", "bourbon", "campari", "centrifuge-clarified grapefruit juice",
                 "champagne acid", "herring cherry liqueur", "clarified lime juice",
                 "clarified lime juice or champagne acid", "cognac", "cointreau",
                 "cranberry juice", "creme de violette", "curacao", "demerara syrup",
                 "dolin dry vermouth", "drambuie", "dry vermouth", "egg white",
                 "fernet branca", "gin", "grapefruit juice", "green chartreuse",
                 "grenadine", "heavy cream", "hellfire bitters", "honey syrup",
                 "lairds applejack bottled in bond", "lemon juice", "lillet blanc",
                 "lime juice", "luxardo maraschino", "old tom gin", "orange bitters",
                 "orange juice", "peychaud's bitters", "pisco", "plymouth gin",
                 "quinine simple syrup", "raspberry syrup", "rye", "saline", "saline solution",
                 "scotch", "simple syrup", "sloe gin", "sugared -proof rum",
                 "sugared -proof rye", "sweet vermouth", "water", "white creme de cacao",
                 "white mezcal", "white rum", "yellow chartreuse")

# Convert the ingredients to a data frame
df <- data.frame(ingredients)

# Define a vector of brand names and processing methods to remove
to_remove <- c("absolut", "centrifuge", "bottled in bond", "agar-", "proof", "lairds", "mechanical separation", "aerated", "- ", '"lime juice or champagne acid"', "old tom ", "dolin", "sugared - ", "sugared -", "peychaud's ", "plymouth " , "luxardo " ,"clarified" , "- ", "sugared" , "hellfire", "blanco" , "herring ")

# Remove brands and processing methods from ingredients
cleaned_ingredients <- df$ingredients

for (pattern in to_remove) {
  cleaned_ingredients <- gsub(pattern, "", cleaned_ingredients)
}

# Trim whitespaces
cleaned_ingredients <- trimws(cleaned_ingredients)

# Remove empty strings
cleaned_ingredients <- cleaned_ingredients[cleaned_ingredients != ""]

# Remove the specified ingredient "bitters"
cleaned_ingredients <- cleaned_ingredients[!grepl("bitters", cleaned_ingredients)]
# Remove the specified ingredient "lime juice or champagne acid"
cleaned_ingredients <- cleaned_ingredients[!grepl("lime juice or champagne acid", cleaned_ingredients)]

# Get unique ingredients
unique_ingredients <- unique(cleaned_ingredients)

# Display unique ingredients
print("\nUnique Ingredients:")
print(unique_ingredients)

#If the brands wouldn't matter nor processes 46 ingredients would be enough. But as it can't be consulted on with the cocktail bar owner I am keeping those in the later data processing.


```

## 5. What are the top 10 ingredients?

```{r}
#5. TOP10 ingredient by count of usage
# Set the width for the console output
options(width = 120)

# Count the frequency of each ingredient
ingredient_counts <- cocktail_data %>%
  count(ingredients, sort = TRUE)

# Create ranks for each ingredient
ingredient_counts <- ingredient_counts %>%
  mutate(rank = dense_rank(desc(n)))

# Display the top 10 ingredients
top_10_ingredients <- ingredient_counts %>%
  filter(rank <= 10)

print("\nTop 10 Ingredients:")
print(top_10_ingredients)

# Display all ingredients with ranks
all_ingredients <- ingredient_counts

print("\nAll Ingredients:")
print(all_ingredients, n = 25)




```

## 6. Which cocktail(s) has/have the most ingredients?

```{r}
# Count the number of ingredients in each cocktail
cocktail_ingredient_counts <- cocktail_data %>%
  group_by(name) %>%
  summarise(ingredient_count = n_distinct(ingredients))

# Find the maximum number of ingredients
max_ingredient_count <- max(cocktail_ingredient_counts$ingredient_count)

# Filter cocktails with the maximum number of ingredients
cocktails_with_most_ingredients <- cocktail_ingredient_counts %>%
  filter(ingredient_count == max_ingredient_count) %>%
  left_join(cocktail_data, by = "name") %>%
  select(name, ingredients)

# Display cocktails with the most ingredients
cat("\nCocktail(s) with the Most Ingredients:\n")

if (nrow(cocktails_with_most_ingredients) == 0) {
  cat("No cocktails found.\n")
} else {
  print(cocktails_with_most_ingredients, n=60)
}

```

## 7. How many ingredients appear in only one cocktail (rare ingredient)?

```{r}

# Exclude specific cocktail and ingredient
excluded_cocktail <- "Carbonated Negroni"
excluded_ingredient <- "clarified lime juice or champagne acid"

# Count the number of cocktails each ingredient appears in
ingredient_counts <- cocktail_data %>%
  group_by(ingredients) %>%
  summarise(cocktail_count = n_distinct(name))

# Filter for ingredients that appear in only one cocktail
unique_ingredients <- ingredient_counts %>%
  filter(cocktail_count == 1) %>%
  left_join(cocktail_data, by = "ingredients") %>%
  group_by(name) %>%
  filter(n() > 0) %>%
  select(name, ingredients)

# Exclude specific cocktail and ingredient
unique_ingredients <- unique_ingredients %>%
  filter(!(name == excluded_cocktail & ingredients == excluded_ingredient))

# Replace specific ingredient name
replacement_ingredient <- "cherry herring"
new_ingredient <- "cherry liqueur"

unique_ingredients <- unique_ingredients %>%
  mutate(ingredients = ifelse(ingredients == replacement_ingredient, new_ingredient, ingredients))

# Display cocktails with at least one unique ingredient
cat("\nCocktails with At Least One Unique Ingredient (cherry herring replaced with cherry liqueur):\n")

if (nrow(unique_ingredients) == 0) {
  cat("No cocktails with at least one unique ingredient found.\n")
} else {

  print(unique_ingredients, n=26)
}

#25 Coctails contain a unique ingredient.
```


## 8. Which cocktail has an ingredient that is only used in one cocktail?

```{r}


cocktail_counts <- cocktail_data %>%
  group_by(name) %>%
  summarise(ingredient_count = n_distinct(ingredients))

# Filter cocktails with only one ingredient
rare_cocktails <- cocktail_counts %>%
  filter(ingredient_count == 1)

# Display rare cocktails and their ingredients
cat("\nRare Cocktails:\n")

if (nrow(rare_cocktails) == 0) {
  cat("No cocktails with only one ingredient found.\n")
} else {
  print(rare_cocktails)
  
  # Get the rare ingredients and their corresponding cocktails
  rare_ingredients <- rare_cocktails %>%
    left_join(cocktail_data, by = "name") %>%
    select(name, ingredients) %>%
    distinct()

  # Display rare ingredients and their corresponding cocktails
  cat("\nRare Ingredients and Cocktails:\n")
  print(rare_ingredients)
}
```


## 9. What are the cocktails without rare ingredients?

```{r}

# Find all unique cocktails
all_cocktails <- cocktail_data %>%
  distinct(name, .keep_all = TRUE) %>%
  select(name)

# Find cocktails without rare ingredients
cocktails_without_rare_ingredients <- anti_join(all_cocktails, unique_ingredients, by = "name")

# Display cocktails without rare ingredients
cat("\nCocktails Without Rare Ingredients:\n")

if (nrow(cocktails_without_rare_ingredients) == 0) {
  cat("No cocktails without rare ingredients found.\n")
} else {
  print(cocktails_without_rare_ingredients, n = 50)
}


```


## 10. Create a cheat sheet for the bartender!
```{r}
# Define corrections
corrections <- c("Manhatan (Boubon, 45% abv)" = "Manhattan (Bourbon, 45% abv)",
                  "Manhatan (Rye, 50% abv)" = "Manhattan (Rye, 50% abv)",
                  "Chartruth" = "Chartreuse")

# Apply corrections to the names
cocktail_data <- cocktail_data %>%
  mutate(name = ifelse(name %in% names(corrections), corrections[name], name))

# Sort the names alphabetically
sorted_cocktail_data <- cocktail_data %>%
  arrange(name)

# Create a matrix that shows all cocktail names as rows and all ingredients as columns
cheat_sheet <- sorted_cocktail_data %>%
  distinct(name, ingredients) %>%
  group_by(name) %>%
  mutate(value = "X") %>%
  pivot_wider(names_from = ingredients, values_from = value, values_fill = "")

# Export the cheat sheet to a CSV file
write.csv(cheat_sheet, file = "cheat_sheet.csv", row.names = FALSE)

# Display the first 3 sorted and corrected cocktails in a tabular format
cat("First 3 Sorted and Corrected Cocktails:\n")
print(head(cheat_sheet, 3))


```

