---
title: 'Assignment 2: Data visualization'
author: "Tamas Nagy"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)
```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

### Read tidytuesday data

```{r}
tt_expedition = tt_load("2020-09-22")
tt_expedition = tt_expedition["expeditions"]
tt_expedition = as.data.frame(tt_expedition)
```

### Wrangle data

```{r}
find_most_frequent_peak = tt_expedition %>% 
                            count(expeditions.peak_name) %>%
                            arrange(desc(n))

find_most_frequent_peak = find_most_frequent_peak[1:15, 1]

is_most_frequent = tt_expedition[, 3] %in% find_most_frequent_peak

expedition_plot_data = tt_expedition[is_most_frequent, ]

expedition_plot_data = expedition_plot_data %>%
                        count(expeditions.season, expeditions.peak_name) %>%
                          arrange(desc(n))

```

### Plotting 

```{r}
p1 = ggplot(expedition_plot_data, aes(x = n, y = expeditions.peak_name, fill = expeditions.season)) +
      geom_bar(position = "stack", stat = "identity")

p1

```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

### Read tidytuesday data

```{r, phd_data}
tt_phd = tt_load("2019-02-19")
tt_phd = tt_phd["phd_by_field"]
tt_phd = as.data.frame(tt_phd)
```

### Data wrangling

```{r, phd_wrangling}
phd_plot_data = tt_phd %>% 
                  count(phd_by_field.year, phd_by_field.broad_field)
```

### Plotting
```{r, phd_plotting}
p2 = ggplot(phd_plot_data, aes(x = phd_by_field.year, y = n, colour = phd_by_field.broad_field)) +
        geom_line()

p2
```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.


### Read data from tidytuesday project

```{r, commute_data}
tt_commute = tt_load("2019-11-05")["commute"]
tt_commute = as.data.frame(tt_commute)
```

### Data wrangling

```{r, commute_wrangling}
commute_plot_data = tt_commute %>%
                      count(commute.state_abb, commute.mode, commute.state_region) %>%
                      pivot_wider(names_from = commute.mode, values_from = n)
                      
```

```{r, commute_plotting}
p3 = ggplot(commute_plot_data, aes(x = Walk, y = Bike, colour = commute.state_region)) + 
      geom_point() +
      scale_x_log10() +
      scale_y_log10()
p3                      
```