---
title: 'Assignment 2: Data visualization'
author: "Tamas Nagy"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
# install.packages('tidyverse')
# install.packages('tidytuesdayR')
library(tidyverse)
library(tidytuesdayR)
library(ggplot2)
```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}
install.packages("viridis")
library(viridis)

climbingExpedition = as.data.frame(tt_load("2020-09-22")["expeditions"])

frequentPeaks <- climbingExpedition %>%
  mutate(peakName = fct_lump(expeditions.peak_name, n = 15)) %>%
  filter(peakName != "Other")

ggplot(frequentPeaks, aes(y=reorder(peakName, table(peakName)[peakName]), fill=expeditions.season)) +
  geom_bar() +
  scale_fill_viridis_d() +
  theme_light() +
  xlab("Number of expeditions") +
  ylab("") +
  theme(legend.position = "bottom") +
  ggtitle("The 15 most popular peaks stacked by season of expedition")

```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}
phdByField = as.data.frame(tt_load("2019-02-19")["phd_by_field"])

numberOfPhDsByYearAndField <- phdByField %>% group_by(phd_by_field.year, phd_by_field.broad_field) %>%
  summarize(numberOfPhDs = sum(phd_by_field.n_phds, na.rm = TRUE))

ggplot(numberOfPhDsByYearAndField, aes(x = phd_by_field.year, y = numberOfPhDs, colour = phd_by_field.broad_field)) +
  geom_line(size = 1.2) +
  scale_colour_brewer(palette = "Dark2") +
  theme_minimal() +
  xlab("") +
  ylab("") +
  labs(colour = "Broad field") +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 5)) +
  scale_y_continuous(labels = scales::comma_format(scale = 1)) +
  ggtitle("Number of awarded Ph.D.-s in the US by year")

```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r}
commutes = as.data.frame(tt_load("2019-11-05")["commute"])

commuteTypes <- commutes %>% group_by(commute.state_abb, commute.mode, commute.state_region) %>%
  summarise(commuters = sum(commute.n)) %>%
  pivot_wider(names_from = commute.mode, values_from = commuters)

ggplot(commuteTypes, aes(x = Walk, y = Bike, colour = commute.state_region, label = commute.state_abb)) +
  geom_point(size = 2) +
  theme_light() +
  geom_text(colour = "#000000") +
  scale_x_log10(labels = scales::comma_format(scale = 1)) +
  scale_y_log10(labels = scales::comma_format(scale = 1), limits = c(20, NA)) +
  xlab("Number of ppl biking to work (log N)") +
  ylab("Number of ppl walking to work (log N)") +
  labs(colour = "State region")

```
