---
title: 'Assignment 2: Data visualization'
author: "Dávid Winkler"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)
library(ggplot2)
library(viridis)
library(forcats)
library(dplyr)
library(scales)
library(tidyr)
expeditions <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-22/expeditions.csv') # We read the data
phd_field <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-02-19/phd_by_field.csv")
commute_mode <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-11-05/commute.csv")

```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}
# Filter out the "Other" named peak from the data
filtered_data <- expeditions %>%
  filter(peak_name != "Other")

# Use fct_lump() to get the 15 most frequent peaks
unique_peaks <- filtered_data$peak_name %>%
  fct_lump(n = 15)

# Reorder the factor levels based on the count of expedition IDs
filtered_data <- filtered_data %>%
  mutate(peak_name = fct_reorder(peak_name, expedition_id, .fun = function(x) sum(!is.na(x)))) %>%
  filter(peak_name %in% unique_peaks)

# Create the bar plot with flipped coordinates
ggplot(filtered_data, aes(x = peak_name, fill = season)) +
  geom_bar() +
  labs(x="", y = "Number of expeditions") + # Setting labels and title
   ggtitle("The 15 most popular peaks stacked by season of expedition") +
  scale_fill_viridis(discrete = TRUE) + # Setting color
  theme_light() + # Setting light theme
  theme(legend.position = "bottom") + # Move the legends to the bottom of the plot
  coord_flip() # Rotate the plot 
```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below! 

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}
# Aggregate the data by broad_field and year
aggregated_data <- aggregate(n_phds ~ broad_field + year, data = phd_field, FUN = sum)

# Set up the plot
ggplot(data = aggregated_data, aes(x = year, y = n_phds, group = broad_field, color = broad_field)) +
geom_line(linewidth = 1.2) +
scale_color_brewer(palette = "Dark2") + # Setting color
labs(title = "Number of awarded Ph.D.-s in the US by year", # Setting labels and title
      x = "",
      y = "",
      color = "Broad field") +
theme_minimal() + # Setting theme minimal
scale_x_continuous(breaks = pretty_breaks()) + # Setting pretty brakes
scale_y_continuous(labels = comma_format()) # Setting comma format



```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r}
# Filter out rows with missing state_region values
filtered_data2 <- commute_mode %>%
  filter(mode %in% c("Walk", "Bike"))

# Aggregate the data by state
aggregated_data2 <- filtered_data2 %>%
  group_by(state) %>%
  summarize(Walk = sum(n[mode == "Walk"]),
            Bike = sum(n[mode == "Bike"]),
            state_region = first(state_region),
            state_abb = first(state_abb))

# Set up the plot
ggplot(data = aggregated_data2, aes(x = Walk, y = Bike, color = state_region, label = state_abb)) +
  geom_point(size = 2) +
  geom_text(color = "Black", check_overlap = TRUE) + # Disable overleaping of state_abbs (monitor size matters)
  labs(title = "Title number of people walking vs. biking to work in each USA state", # Setting title and labels
       x = "Number of ppl biking to work (log N)",
       y = "Number of ppl walking to work (log N)",
       color = "State region") + 
  scale_x_log10(labels = comma_format()) + # Log transformations and coma formats
  scale_y_log10(labels = comma_format()) +
  theme_light() # Setting theme light

```
