---
title: 'Assignment 2: Data visualization'
author: "Tamas Nagy"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)
library(scales)
library(ggplot2)
```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}
expeditions <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-22/expeditions.csv")

expeditions_clean <- expeditions %>%
  select(peak_name, season) %>%
  mutate(peak_name = fct_lump_n(peak_name, 15)) %>%
  filter(peak_name != "Other")

expeditions_groupby <- expeditions_clean %>%
  count(peak_name, season)

ggplot(expeditions_groupby, aes(x = n, y = fct_reorder(peak_name, n, .fun = sum), fill = season)) +
  geom_col() +
  theme_light() +
  scale_fill_viridis_d() +
  labs(title = "The 15 most popular peaks stacked by season of expedition", x = "Number of expeditions", y = NULL) +
  theme(
    axis.title.y = element_blank(),
    legend.position = "bottom"
  )
```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}
phd_by_field <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-02-19/phd_by_field.csv")

phd_by_field_groupby <- phd_by_field %>%
  filter(!is.na(n_phds)) %>%
  group_by(broad_field, year) %>%
  summarize(Count = sum(n_phds))

ggplot(phd_by_field_groupby, aes(x = year, y = Count, color = broad_field)) +
  geom_line(size = 1.2) +
  scale_x_continuous(breaks = pretty_breaks(6)) +
  scale_y_continuous(label = comma_format()) +
  scale_colour_brewer(palette = "Dark2", name = "Broad field") +
  theme_minimal() +
  labs(title = "Number of awarded phd-s by year")
```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r}
commute <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-11-05/commute.csv")

commute_pivoted <- commute %>%
  group_by(mode, state_abb, state_region) %>%
  summarize(Count = sum(n)) %>%
  pivot_wider(names_from = "mode", values_from = "Count")

ggplot(commute_pivoted, aes(x = Walk, y = Bike, color = state_region)) +
  geom_point(size = 2) +
  geom_text(aes(label = state_abb), colour = "black") +
  scale_x_log10(label = comma_format()) +
  scale_y_log10(limits = c(100, 150000), label = comma_format()) +
  theme_light() +
  labs(
    x = "Number of people walking to work (log N)",
    y = "Number of people biking to work (log N)",
    title = "Number of people walking vs. biking to work in each USA state",
    color = "State region")
```
