---
title: 'Assignment 2: Data visualization'
author: "Bela Szlovicsak"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)
library(scales)
```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}
expeditions <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-22/expeditions.csv')
expeditions %>% 
  select(peak_name, season) %>% 
  mutate(peak_name = fct_lump_n(peak_name, 15)) %>% 
  filter(peak_name != "Other") %>% 
  ggplot(aes(x =
      fct_rev(fct_infreq(peak_name)),
      fill = season
  )) +
  geom_bar(position = "stack") +
  coord_flip() +
  theme_light() +
  scale_fill_viridis_d() +
  theme(
    axis.title.y = element_blank(),
    legend.position = "bottom"
  ) +
  labs(y = "Number of expeditions",
       title = "The 15 most popular peaks stacked by season of expedition")
  

```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}
phd_by_field <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-02-19/phd_by_field.csv")

phd_by_field %>%
  filter(!is.na(n_phds)) %>% 
  group_by(broad_field, year) %>%
  summarize(n = sum(n_phds)) %>%
  ggplot(aes(
      year,
      n,
      color = broad_field
  )) +
  geom_line(size = 1.2) +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal() +
  theme(axis.title.y = element_blank(),
        axis.title.x = element_blank()) +
  labs(color = "Broad field",
       title = "Number of awarded Ph.D.-s in the US by Year") +
  scale_x_continuous(breaks = pretty_breaks()) +
  scale_y_continuous(labels = label_comma())
  

```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r}
commute_mode <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-11-05/commute.csv")

commute_mode %>% 
  group_by(mode, state_abb, state_region) %>% 
  summarise(n = sum(n)) %>% 
  pivot_wider(names_from = mode,
              values_from = n,
  ) %>% 
  ggplot(
    aes(
      Walk,
      Bike,
      color = state_region
    )
  ) +
  geom_point(size = 2) +
  geom_text(aes(label = state_abb), 
            color = "black",
            check_overlap = TRUE) +
  scale_x_log10(labels = label_comma()) +
  scale_y_log10(labels = label_comma(),
                limits = c(20, 150000)) +
  theme_light() +
  labs(x = "Number of ppl walking to work (log N)",
       y = "Number of ppl biking to work (log N)",
       title = "Number of people walking vs. biking to work in each USA state",
       color = "State region")


```
