---
title: 'Assignment 2: Data visualization'
author: "Tamas Nagy"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)
library(viridis)
```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}
tt_expedition <- tt_load("2020-09-22")
tt_expedition <- tt_expedition$expeditions
```


```{r}

#tt_expedition$season <- as_factor(tt_expedition$season)
#tt_expedition$peak_name <- as_factor(tt_expedition$peak_name)

max15_peak <- tt_expedition %>%
    group_by(season) %>%
     count(peak_name) %>%
     arrange(desc(n))

max15_peak$peak_name <- fct_lump(max15_peak$peak_name, 15)
max15_peak$peak_name <- fct_reorder(max15_peak$peak_name, max15_peak$n, sum)

max15_peak <- max15_peak %>%
  filter(peak_name != "Other")


plot1 <- max15_peak %>%
  ggplot() + 
  aes(x = peak_name, y = n, fill = season) +
  geom_bar(stat = "identity") + 
  scale_fill_viridis() +
  ggtitle("The 15 most popular peaks stacked by season of expedition") + 
  ylab("") +
  xlab("Number of expeditions") +
  guides(fill=guide_legend(title="season")) +
  theme(legend.position = "bottom") +
  theme_light() +
  scale_y_continuous() +
  scale_x_discrete() +
  coord_flp

#i can't understant the error here, i assume it's because of the factor and identity statistics mess up 
#but i couldn't resolve this anyhow with the scales or ordering or even transfroming to characters

#Error: Discrete value supplied to continuous scale

```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}
tt_phd <- tt_load("2019-02-19")
tt_phd <- tt_phd$phd_by_field
```

```{r}
library(scales)

phd_plot = tt_phd %>% 
    group_by(year, broad_field) %>%
    summarise(doctors = sum(n_phds, na.rm = TRUE))

plot2 = ggplot(phd_plot, aes(x = year, y = doctors, colour = broad_field)) +
        geom_line(size = 1.2)+
        ggtitle("Number of awarded Ph.D.-s in the US by year") +
        xlab("") +
        ylab("") +
        labs(colour = "Broad field") +
        scale_y_continuous(labels = comma_format(scale = 1)) +
        scale_x_continuous(breaks = pretty_breaks(5)) +
        scale_colour_brewer(palette = "Dark2") +
        theme_minimal()
```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r}
tt_commute <- tt_load("2019-11-05")
tt_commute <- tt_commute$commute

```